---
title: "Getting Started with FastImputation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with FastImputation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
library(FastImputation)
```

## Overview

FastImputation fills in missing values using the conditional multivariate
normal (MVN) distribution — the same approach used by
[Amelia](https://gking.harvard.edu/amelia), but optimised for speed when
imputing a single row at a time. It is well-suited to pre-processing steps
before running row-wise prediction pipelines.

The package provides three primary interfaces:

1. `train_fast_imputation()` + `fast_imputation()` — standalone two-step API.
2. `step_fast_imputation()` — drop-in step for
   [recipes](https://recipes.tidymodels.org) / tidymodels pipelines.

## Basic usage

```{r basic}
data(FI_train)
data(FI_test)

# Step 1: learn from training data
# FI_train has an ID column and a categorical column that must be declared
patterns <- train_fast_imputation(
  FI_train,
  ignore_cols = user_id_1,
  categorical = categorical_9
)

# Step 2: impute a data frame with missing values
imputed <- fast_imputation(FI_test, patterns)
head(imputed)
```

The result is a data frame (or tibble) with the same columns as the input,
but with `NA` values replaced by the conditional MVN estimate.

## Pipe-friendly interface

`fast_imputation()` puts the data as its first argument, so it works naturally
with the base R pipe:

```{r pipe, eval = FALSE}
FI_test |> fast_imputation(patterns)
```

## Bounded variables

Some variables are physically constrained to a range — for example, age
cannot be negative, or a probability must be between 0 and 1.  Pass a
`constraints` list to preserve these bounds:

```{r bounded}
patterns_bounded <- train_fast_imputation(
  FI_train,
  constraints = list(
    bounded_below_2           = list(lower = 0),
    bounded_above_5           = list(upper = 0),
    bounded_above_and_below_6 = list(lower = 0, upper = 1)
  ),
  ignore_cols = user_id_1,
  categorical = categorical_9
)
imputed_bounded <- fast_imputation(FI_test, patterns_bounded)
```

* `lower` — the variable is bounded from below (e.g. `lower = 0`).
* `upper` — the variable is bounded from above.
* Both `lower` and `upper` — the variable is bounded on an interval.

Internally, bounded variables are transformed to the real line before the
MVN estimation (log-transform for one-sided bounds; inverse-normal CDF for
two-sided bounds), and then transformed back after imputation.

## Categorical variables

Categorical columns are one-hot encoded internally so they integrate
seamlessly with the MVN model.  Use the `categorical` argument with any
[tidyselect](https://tidyselect.r-lib.org) expression:

```{r categorical}
patterns_cat <- train_fast_imputation(
  FI_train,
  ignore_cols = user_id_1,
  categorical = categorical_9
)
imputed_cat <- fast_imputation(FI_test, patterns_cat)
head(imputed_cat$categorical_9)
```

After imputation, the categorical column is returned as a factor with the
same levels as in the training data.

## Ignoring columns (e.g. IDs)

Pass ID columns or other columns that should be carried through unchanged
via `ignore_cols`.  Any tidyselect expression is accepted:

```{r ignore}
patterns_id <- train_fast_imputation(
  FI_train,
  ignore_cols = user_id_1,
  categorical = categorical_9
)
```

You can combine all options:

```{r combined}
patterns <- train_fast_imputation(
  FI_train,
  constraints = list(
    bounded_below_2           = list(lower = 0),
    bounded_above_and_below_6 = list(lower = 0, upper = 1)
  ),
  ignore_cols = user_id_1,
  categorical = categorical_9
)
```

## Inspecting a patterns object

`print()` gives a quick summary:

```{r print}
patterns
```

`tidy()` (from the [generics](https://generics.r-lib.org) package) returns
a tibble with one row per variable:

```{r tidy}
generics::tidy(patterns)
```

## tidymodels / recipes integration

`step_fast_imputation()` lets you use FastImputation inside a
[recipes](https://recipes.tidymodels.org) pre-processing pipeline:

```{r recipes, eval = FALSE}
library(recipes)

rec <- recipe(~ ., data = FI_train) |>
  step_fast_imputation(
    everything(),
    constraints = list(
      bounded_below_2           = list(lower = 0),
      bounded_above_and_below_6 = list(lower = 0, upper = 1)
    ),
    ignore_cols = user_id_1,
    categorical = categorical_9
  )

trained_rec <- prep(rec, training = FI_train)
baked       <- bake(trained_rec, new_data = FI_test)
head(baked)
```

The step integrates with `tidy()` for inspection:

```{r recipes_tidy, eval = FALSE}
tidy(trained_rec, number = 1)
```

## Accuracy comparison

```{r accuracy}
data(FI_true)

patterns_full <- train_fast_imputation(
  FI_train,
  constraints = list(
    bounded_below_2           = list(lower = 0),
    bounded_above_5           = list(upper = 0),
    bounded_above_and_below_6 = list(lower = 0, upper = 1)
  ),
  ignore_cols = user_id_1,
  categorical = categorical_9
)

imputed_full <- fast_imputation(FI_test, patterns_full)

# Compare continuous imputation accuracy
continuous_missing <- is.na(FI_test[, 2:8])
imputed_vals       <- imputed_full[, 2:8][continuous_missing]
true_vals          <- FI_true[, 2:8][continuous_missing]

rmse <- sqrt(median((imputed_vals - true_vals)^2))
cat("FastImputation RMSE:", round(rmse, 4), "\n")

# Baseline: column means
col_mean_imputed <- FI_test[, 2:8]
for (j in seq_len(ncol(col_mean_imputed))) {
  col_mean_imputed[is.na(col_mean_imputed[, j]), j] <-
    mean(col_mean_imputed[, j], na.rm = TRUE)
}
rmse_means <- sqrt(median((col_mean_imputed[continuous_missing] - true_vals)^2))
cat("Column-means RMSE: ", round(rmse_means, 4), "\n")
```

FastImputation consistently achieves lower error than simple column-mean
imputation by leveraging correlations between variables.
